### 2015-02-09 doc/thesis/system.tex
Minor improvements to the text about case studies
Adding regular table types to the text

\begin{figure}[!ht]
\textbf{Type Language}\\
\dstart
$$
\begin{array}{rlr}
t ::= & & \textsc{first-level types:}\\
& \;\; l & \textit{literal types}\\
& | \; b & \textit{base types}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; t \cup t & \textit{disjoint union types}\\
& | \; s \rightarrow s & \textit{function types}\\
& | \; \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{closed|open|unique} & \textit{table types}\\
& | \; x & \textit{type variables}\\
& | \; \mu x.t & \textit{recursive types}\\
& | \; \phi(t,t) & \textit{filter types}\\
& | \; \pi_{i}^{x} & \textit{projection types}\\
%\multicolumn{3}{c}{}\\
l ::= & & \textsc{{\small literal types:}}\\
& \;\; \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} &\\
%\multicolumn{3}{c}{}\\
b ::= & & \textsc{{\small base types:}}\\
& \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String &\\
%\multicolumn{3}{c}{}\\
k ::= & & \textsc{{\small key types:}}\\
& \;\; l \; | \; b \; | \; \Value &\\
%\multicolumn{3}{c}{}\\
v ::= & & \textsc{{\small value types:}}\\
& \;\; t \; | \; \Const \; t &\\ 
%\multicolumn{3}{c}{}\\
s ::= & & \textsc{second-level types:}\\
& \;\; p & \textit{tuple types}\\
& | \; s \sqcup s & \textit{unions of tuple types}\\
%\multicolumn{3}{c}{}\\
p ::= & & \textsc{{\small tuple types:}}\\
& \;\; \Void & \textit{void type}\\
& | \; t* & \textit{variadic types}\\
& | \; t \times p & \textit{pair types}
\end{array}
$$
\dend
\caption{The abstract syntax of Typed Lua types}
\label{fig:typelang}
\end{figure}

@@

Back to first-level types, table types represent the various forms
that Lua tables can take.
The syntactical form of table types is $\{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n} \}_{tag}$,
where each $k_{i}$ represents the type of a table key,
and each $v_{i}$ represents the type of the value that table keys of type $k_{i}$ map to.
Key types can only be literal types, base types, or the top type.
We made this restriction to the type of the keys because the statistics
that we discussed in Section \ref{sec:statistics} showed that most
of the tables are records, lists, and hashes.
The type $\Value$ is an option when we need a loose table type.
For instance, $\{\Value:\Value\}_{closed}$ represents the type of a
table in which both indices and values can have any type.
Value types can be any first-level type, and can optionally include
the $\Const$ type to denote immutable values.

We also use the tags \emph{closed}, \emph{open}, and \emph{unique}
to classify table types.
The tag \emph{unique} represents tables with no keys that do not
inhabit one of the table's key types, and with no alias.
In particular, the type of the table constructor has this tag.
The tag \emph{open} represents \emph{unique} table types that
have at least one alias.
The tag \emph{closed} represents table types that do not provide
any guarantees about keys with types not listed in the table type.
In particular, in the concrete syntax, type annotations, interface
declarations, and userdata declarations always describe \emph{closed} table types.
In the next sections we explain in more detail why we need
different table types.

@@

Recursive types have the form $\mu x.t$,
where $t$ is a first-level type that $x$ represents.
For instance, $\mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed}$
is a type for singly-linked lists of integers.
In Section \ref{sec:alias} we mentioned that we can use the following
interface declaration as an alias to this type:
\begin{verbatim}
    local interface Element
      info:integer
      next:Element?
    end
\end{verbatim}

@@

Back to the subtyping rules between first-level types,
the subtyping rule between \emph{closed} table types resembles the
standard subtyping rule between records:
\[
\begin{array}{c}
\mylabel{S-TABLE1}\\
\dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
       \senv \vdash k_{j} \subtype k_{i}' \;\;\;
       \senv \vdash k_{i}' \subtype k_{j} \;\;\;
       \senv \vdash v_{j} \subtype_{c} v_{i}'}
      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{closed} \subtype \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{closed}} \; m \ge n
\end{array}
\]

The rule \textsc{S-TABLE1} allows width subtyping between \emph{closed}
table types, and introduces the auxiliary relation $\subtype_{c}$ to
handle depth subtyping on the type of the values stored in the table fields.
We need an auxiliary relation because the subtyping of the
type of the values stored in the table fields changes according to
the tags of the table types.
We define the relation $\subtype_{c}$ as follows:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FIELD1}\\
\dfrac{\senv \vdash v_{1} \subtype v_{2} \;\;\;
       \senv \vdash v_{2} \subtype v_{1}}
      {\senv \vdash v_{1} \subtype_{c} v_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD2}\\
\dfrac{\senv \vdash v_{1} \subtype v_{2}}
      {\senv \vdash \Const \; v_{1} \subtype_{c} \Const \; v_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD3}\\
\dfrac{\senv \vdash v_{1} \subtype v_{2}}
      {\senv \vdash v_{1} \subtype_{c} \Const \; v_{2}}
\end{array}
\end{array}
\]

These rules allow closed table types to have depth subtyping on $\Const$ fields.
The rule \textsc{S-FIELD1} defines that mutable fields are invariant,
while the rule \textsc{S-FIELD2} defines that immutable fields are covariant.
The rule \textsc{S-FIELD3} defines that it is safe to promote fields
from mutable to immutable.
We do not include a rule that allows promoting fields from immutable
to mutable because this would be unsafe due to variance.

There is a limitation on \emph{closed} table types that led us to
introduce \emph{open} and \emph{unique} table types.
If the table constructor had a \emph{closed} table type, then
programmers would not be able to use it to initialize a variable with
a table type that describes a more general type.
For instance,
\begin{verbatim}
    local t:{"x":integer, "y":integer?} = { x = 1, y = 2 }
\end{verbatim}
would not type check, as the type of the table constructor would not
be a subtype of the type in the annotation.
More precisely,
\[
\{``x":1, ``y":2\}_{closed} \not\subtype \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
\]

Simply promoting the type of each table value to its supertype would
not overcome this limitation, as it still would give to the table constructor
a closed table type without covariant mutable fields.
Thus, programmers would not be able to use the table constructor to
initialize a variable with a table type that includes an optional field.
Using the previous example,
\begin{align*}
& \{``x":\Integer, ``y":\Integer\}_{closed} \not\subtype \\
& \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
\end{align*}

We introduced \emph{unique} table types to avoid this limitation,
as they represent the type of tables with no keys that do not
inhabit one of the table's key types, and with no alias.
In particular, this is the case of the table constructor.
The following subtyping rule defines the subtyping relation among
\emph{unique} table types and \emph{closed} table types:
\[
\begin{array}{c}
\mylabel{S-TABLE2}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..m \; \forall j \in 1..n \;
       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash v_{i} \subtype_{u} v_{j}' \\
       \forall j \in 1..n \; \not\exists i \in 1..m \;
       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash \Nil \subtype_{o} v_{j}'
       \end{array}}
      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{unique} \subtype
                    \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{closed}}
\end{array}
\]

The rule \textsc{S-TABLE2} allows width subtyping and covariant keys.
It allows covariant keys because we also want to use \emph{unique}
table types as a way to join table fields that inhabit \emph{closed} table types.
For instance, we want to use the table constructor to initialize
a variable with a table type that describes a hash.

@@

Using \emph{unique} table types to represent the type of the table
constructor allows our type system to type check the previous example.
More precisely,
\[
\{``x":1, ``y":2\}_{unique} \subtype \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
\]

Even though we allow width subtyping between \emph{unique} and \emph{closed}
table types, we do not allow it among \emph{unique} and \emph{open} or
\emph{unique} table types because it would violate our definition of
\emph{unique} and \emph{open} table types:
\[
\begin{array}{c}
\mylabel{S-TABLE3}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..m \\
       \exists j \in 1..n \;
       \senv \vdash k_{i} \subtype k_{j}' \land \senv \vdash v_{i} \subtype_{u} v_{j}' \\
       \forall j \in 1..n \; \not\exists i \in 1..m \;
       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash \Nil \subtype_{o} v_{j}'
       \end{array}}
      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{unique} \subtype
                    \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{open|unique}}
\end{array}
\]

The rules that handle subtyping among \emph{open} table types and
\emph{open} or \emph{closed} table types are similar to the rules
that handle subtyping among \emph{unique} table types and other table types:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-TABLE4}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..m \; \forall j \in 1..n \;
       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash v_{i} \subtype_{c} v_{j}' \\
       \forall j \in 1..n \; \not\exists i \in 1..m \;
       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash \Nil \subtype_{o} v_{j}'
       \end{array}}
      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{open} \subtype
                    \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{closed}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-TABLE5}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..m \\
       \exists j \in 1..n \;
       \senv \vdash k_{i} \subtype k_{j}' \land \senv \vdash v_{i} \subtype_{c} v_{j}' \\
       \forall j \in 1..n \; \not\exists i \in 1..m \;
       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash \Nil \subtype_{o} v_{j}'
       \end{array}}
      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{open} \subtype
                    \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{open}}
\end{array}
\end{array}
\]

The rule \textsc{S-TABLE4} allows width subtyping, while
the rule \textsc{S-TABLE5} does not allow width subtyping.
Both rules allow joining fields plus omitting optional fields.
Both rules use $\subtype_{c}$ to allow depth on $\Const$ fields only.

In the next section we will show in more detail how our type system
handles \emph{open} and \emph{unique} table types to allow the refinement of
table types.

@@

As an example, the rule \textsc{S-UNFOLDR} allows our type system to
type check the function \texttt{insert} from Section \ref{sec:alias}:
\begin{verbatim}
    local function insert (e:Element?, v:integer):Element
      return { info = v, next = e }
    end
\end{verbatim}
that is, the type checker uses the rule \textsc{S-UNFOLDR} to verify whether
the type of the table constructor is a subtype of \texttt{Element}:
\begin{align*}
\{ & ``info":\Integer, \\
   & ``next":\mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed} \cup \Nil \}_{unique} \subtype \\
& \mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed}
\end{align*}

@@

Our reduced core also splits function application, method application, and
the vararg expression (${...}$) into different syntactic categories.
It uses $e(el)_{s}$ to denote function applications that return no value
because they appear as statements,
$e(el)_{e}$ to denote function applications that return only one value,
and $e(el)_{m}$ to denote function applications that return multiple values.
It uses the same categories for method applications, but only ${...}_{e}$
and ${...}_{m}$, as the vararg expression cannot appear as a statement.

We also include two kinds of type cast in our core language:
the expression ${<}t{>} \;n$ and the left-hand value $n[c] \; {<}t{>}$.
We will show that the first helps allowing safe aliasing between
table types, while the second allows the refinement of table types.
Due to the refinement of table types we also split variable names
into two categories: $n_{e}$ when they appear as expressions and
$n_{l}$ when they appear as left-hand values.

We use three different typing relations to present our typing rules
as a deduction system for these typing relations.
We use the relation $\env_{1}, \penv \vdash s, \env_{2}$ for typing
statements.
This relation means that given the type environment $\env_{1}$
and the projection environment $\penv$,
typing the statement $s$ produces the new type environment $\env_{2}$.
We use the relation $\env_{1}, \penv \vdash e : t, \env_{2}$ for
typing expressions.
This relation means that given the type environment $\env_{1}$
and the projection environment $\penv$,
the expression $e$ has type $t$ and produces the new type environment $\env_{2}$.
We use the relation $\env_{1}, \penv \vdash el : r_{1}, \env_{2}, (x,r_{2})$
for typing expression lists.
This relation means that given type type environment $\env_{1}$
and the projection environment $\penv$,
the expression list $el$ has type $r_{1}$ and produces the new type
environment $\env_{2}$ and produces the pair $(x,r_{2})$.
We need three typing relations because statements do not have types
and expression lists can result in a union of tuple types that
should be introduced in the projection environment.
We use $\env_{1}, \penv \vdash el : r, \env_{2}$ when the expression
list does not result in a union of tuple types.
We need two environments because $\env$ maps variables to types,
while $\penv$ maps projection variables to unions of tuple types.
We use $\env_{1}[n \mapsto t]$ for meaning that we extended the environment
$\env_{1}$ with the variable $n$ that maps to type $t$.
We use $\penv[x \mapsto r]$ for meaning that we extended the environment
$\penv$ with the projection variable $x$ that maps to the union of tuple types $r$.

Lua has multiple assignments, and the rule \textsc{T-ASSIGNMENT} uses
tuple types along with consistent-subtyping to type check this Lua feature:
\[
\begin{array}{c}
\mylabel{T-ASSIGNMENT}\\
\begin{array}{c}
\dfrac{\env_{1}, \penv \vdash el:r_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash \vec{l}:r_{2}, \env_{3} \;\;\;
       r_{1} \lesssim r_{2}}
      {\env_{1}, \penv \vdash \vec{l} = el,\env_{3}}
\end{array}
\end{array}
\]

As an example,
\[
x, y = 1, ``foo"
\]
type checks because
$1 \times ``foo" \times \Nil{*} \lesssim \Integer \times \String \times \Value{*}$,
assuming that $x$ and $y$ are in the environment with
types $\Integer$ and $\String$, respectively.
Our type system uses $\Value{*}$ to drop extra values
and $\Nil{*}$ to replace missing values.
In this example, the rules \textsc{T-LHSLIST} and \textsc{T-EXPLIST2}
are the rules that handle these features.

We define the rule \textsc{T-LHSLIST} as follows:
\[
\begin{array}{c}
\mylabel{T-LHSLIST}\\
\dfrac{\env, \penv \vdash l_{k}:t_{k}, \env_{k} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
       n = |\;\vec{l}\;|}
      {\env, \penv \vdash \vec{l}:t_{1} \times ... \times t_{n} \times \Value{*}, \env_{m}}
\end{array}
\]

First, this rule uses the same environment $\env$ to type check each
left-hand side value $l_{k}$, as they can perform different changes
in the environment.
Then, it takes each type $t_{k}$ and builds the tuple type for
the assignment rule.
It places the type $\Value{*}$ in the end to discard extra values.
The rule succeeds if the predicate \emph{merge} can
produce a new environment, which includes all the modifications
of each environment $\env_{k}$.
Intuitively, the predicate \emph{merge} fails when an environment
tries to change the type of a variable towards a type that is not
a subtype of its previous type.
We will show in this section that the refinement of table types can
change the environment, and that is the reason why we need to check
whether the environment is consistent.

>>>>

\begin{figure}[!ht]
\textbf{Type Language}\\
\dstart
$$
\begin{array}{rlr}
t ::= & & \textsc{first-level types:}\\
& \;\; l & \textit{literal types}\\
& | \; b & \textit{base types}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; t \cup t & \textit{disjoint union types}\\
& | \; s \rightarrow s & \textit{function types}\\
& | \; \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{unique|open|closed|regular} & \textit{table types}\\
& | \; x & \textit{type variables}\\
& | \; \mu x.t & \textit{recursive types}\\
& | \; \phi(t,t) & \textit{filter types}\\
& | \; \pi_{i}^{x} & \textit{projection types}\\
%\multicolumn{3}{c}{}\\
l ::= & & \textsc{{\small literal types:}}\\
& \;\; \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} &\\
%\multicolumn{3}{c}{}\\
b ::= & & \textsc{{\small base types:}}\\
& \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String &\\
%\multicolumn{3}{c}{}\\
k ::= & & \textsc{{\small key types:}}\\
& \;\; l \; | \; b \; | \; \Value &\\
%\multicolumn{3}{c}{}\\
v ::= & & \textsc{{\small value types:}}\\
& \;\; t \; | \; \Const \; t &\\ 
%\multicolumn{3}{c}{}\\
s ::= & & \textsc{second-level types:}\\
& \;\; p & \textit{tuple types}\\
& | \; s \sqcup s & \textit{unions of tuple types}\\
%\multicolumn{3}{c}{}\\
p ::= & & \textsc{{\small tuple types:}}\\
& \;\; \Void & \textit{void type}\\
& | \; t* & \textit{variadic types}\\
& | \; t \times p & \textit{pair types}
\end{array}
$$
\dend
\caption{The abstract syntax of Typed Lua types}
\label{fig:typelang}
\end{figure}

@@

Back to first-level types, table types represent the various forms
that Lua tables can take.
The syntactical form of table types is $\{ k_{1}{:}v_{1}, ..., k_{n}{:}v_{n} \}_{tag}$,
where each $k_{i}$ represents the type of a table key,
and each $v_{i}$ represents the type of the value that table keys of type $k_{i}$ map to.
Key types can only be literal types, base types, or the top type.
We made this restriction to the type of the keys because the statistics
that we discussed in Section \ref{sec:statistics} showed that most
of the tables are records, lists, and hashes.
The type $\Value$ is an option when we need a loose table type.
For instance, $\{\Value:\Value\}_{regular}$ represents the type of a
table in which both indices and values can have any type.
Value types can be any first-level type, and can optionally include
the $\Const$ type to denote immutable values.

We also use the tags \emph{unique}, \emph{open}, \emph{closed}, and
\emph{regular} to classify table types.
The tag \emph{unique} represents tables with no keys that do not
inhabit one of the table's key types, and with no alias.
In particular, the type of the table constructor has this tag.
The tag \emph{open} represents \emph{unique} table types that
have at least one alias.
The tag \emph{closed} represents \emph{unique} table types
that may have aliases.
In particular, the type of a class has this tag.
The tag \emph{regular} represents table types that do not provide
any guarantees about keys with types not listed in the table type.
In particular, in the concrete syntax, type annotations, interface
declarations, and userdata declarations always describe \emph{regular} table types.
In the next sections we explain in more detail why we need
different table types.

@@

Recursive types have the form $\mu x.t$,
where $t$ is a first-level type that $x$ represents.
For instance, $\mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{regular}$
is a type for singly-linked lists of integers.
In Section \ref{sec:alias} we mentioned that we can use the following
interface declaration as an alias to this type:
\begin{verbatim}
    local interface Element
      info:integer
      next:Element?
    end
\end{verbatim}

@@

Back to the subtyping rules between first-level types,
the subtyping rule among a \emph{closed} or \emph{regular}
table type and another \emph{regular} table type resembles the
standard subtyping rule between records:
\[
\begin{array}{c}
\mylabel{S-TABLE1}\\
\dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
       \senv \vdash k_{j} \subtype k_{i}' \;\;\;
       \senv \vdash k_{i}' \subtype k_{j} \;\;\;
       \senv \vdash v_{j} \subtype_{c} v_{i}'}
      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{closed|regular} \subtype \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{regular}} \; m \ge n
\end{array}
\]

The rule \textsc{S-TABLE1} allows width subtyping and introduces the
auxiliary relation $\subtype_{c}$ to handle depth subtyping on the
type of the values stored in the table fields.
We need an auxiliary relation because the subtyping of the
type of the values stored in the table fields changes according to
the tags of the table types.
We define the relation $\subtype_{c}$ as follows:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FIELD1}\\
\dfrac{\senv \vdash v_{1} \subtype v_{2} \;\;\;
       \senv \vdash v_{2} \subtype v_{1}}
      {\senv \vdash v_{1} \subtype_{c} v_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD2}\\
\dfrac{\senv \vdash v_{1} \subtype v_{2}}
      {\senv \vdash \Const \; v_{1} \subtype_{c} \Const \; v_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD3}\\
\dfrac{\senv \vdash v_{1} \subtype v_{2}}
      {\senv \vdash v_{1} \subtype_{c} \Const \; v_{2}}
\end{array}
\end{array}
\]

These rules allow depth subtyping on $\Const$ fields.
The rule \textsc{S-FIELD1} defines that mutable fields are invariant,
while the rule \textsc{S-FIELD2} defines that immutable fields are covariant.
The rule \textsc{S-FIELD3} defines that it is safe to promote fields
from mutable to immutable.
We do not include a rule that allows promoting fields from immutable
to mutable because this would be unsafe due to variance.

There is a limitation on \emph{regular} table types that led us to
introduce \emph{open} and \emph{unique} table types.
If the table constructor had a \emph{regular} table type, then
programmers would not be able to use it to initialize a variable with
a table type that describes a more general type.
For instance,
\begin{verbatim}
    local t:{"x":integer, "y":integer?} = { x = 1, y = 2 }
\end{verbatim}
would not type check, as the type of the table constructor would not
be a subtype of the type in the annotation.
More precisely,
\[
\{``x":1, ``y":2\}_{regular} \not\subtype \{``x":\Integer, ``y":\Integer \cup \Nil\}_{regular}
\]

Simply promoting the type of each table value to its supertype would
not overcome this limitation, as it still would give to the table constructor
a regular table type without covariant mutable fields.
Thus, programmers would not be able to use the table constructor to
initialize a variable with a table type that includes an optional field.
Using the previous example,
\begin{align*}
& \{``x":\Integer, ``y":\Integer\}_{regular} \not\subtype \\
& \{``x":\Integer, ``y":\Integer \cup \Nil\}_{regular}
\end{align*}

We introduced \emph{unique} table types to avoid this limitation,
as they represent the type of tables with no keys that do not
inhabit one of the table's key types, and with no alias.
In particular, this is the case of the table constructor.
The following subtyping rule defines the subtyping relation among
\emph{unique} table types and \emph{regular} table types:
\[
\begin{array}{c}
\mylabel{S-TABLE2}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..m \; \forall j \in 1..n \;
       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash v_{i} \subtype_{u} v_{j}' \\
       \forall j \in 1..n \; \not\exists i \in 1..m \;
       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash \Nil \subtype_{o} v_{j}'
       \end{array}}
      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{unique} \subtype
                    \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{regular}}
\end{array}
\]

The rule \textsc{S-TABLE2} allows width subtyping and covariant keys.
It allows covariant keys because we also want to use \emph{unique}
table types as a way to join table fields that inhabit \emph{regular} table types.
For instance, we want to use the table constructor to initialize
a variable with a table type that describes a hash.

@@

Using \emph{unique} table types to represent the type of the table
constructor allows our type system to type check the previous example.
More precisely,
\[
\{``x":1, ``y":2\}_{unique} \subtype \{``x":\Integer, ``y":\Integer \cup \Nil\}_{regular}
\]

Even though we allow width subtyping between \emph{unique} and \emph{regular}
table types, we do not allow it among \emph{unique} and other table types
because it would violate our definition of these other table types:
\[
\begin{array}{c}
\mylabel{S-TABLE3}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..m \\
       \exists j \in 1..n \;
       \senv \vdash k_{i} \subtype k_{j}' \land \senv \vdash v_{i} \subtype_{u} v_{j}' \\
       \forall j \in 1..n \; \not\exists i \in 1..m \;
       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash \Nil \subtype_{o} v_{j}'
       \end{array}}
      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{unique} \subtype
                    \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{unique|open|closed}}
\end{array}
\]

The rule that handles subtyping between \emph{open} and \emph{regular} table
types allows width subtyping:
\[
\begin{array}{c}
\mylabel{S-TABLE4}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..m \; \forall j \in 1..n \;
       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash v_{i} \subtype_{c} v_{j}' \\
       \forall j \in 1..n \; \not\exists i \in 1..m \;
       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash \Nil \subtype_{o} v_{j}'
       \end{array}}
      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{open} \subtype
                    \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{regular}}
\end{array}
\]

However, the rule that handles subtyping among \emph{open} and
\emph{open} or \emph{closed} table types does not allow width subtyping:
\[
\begin{array}{c}
\mylabel{S-TABLE5}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..m \\
       \exists j \in 1..n \;
       \senv \vdash k_{i} \subtype k_{j}' \land \senv \vdash v_{i} \subtype_{c} v_{j}' \\
       \forall j \in 1..n \; \not\exists i \in 1..m \;
       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash \Nil \subtype_{o} v_{j}'
       \end{array}}
      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{open} \subtype
                    \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{open|closed}}
\end{array}
\]

The rules \textsc{S-TABLE4} and \textsc{S-TABLE5} allow joining fields
plus omitting optional fields.
Both rules use $\subtype_{c}$ to allow depth subtyping on $\Const$ fields only.

We introduced \emph{closed} table types because we needed a safe way
to represent the type of classes along with single inheritance.
The rule that handles subtyping between \emph{closed} table types
does not allow width subtyping, joining fields, and omitting fields,
but it allows depth subtyping on $\Const$ fields:
\[
\begin{array}{c}
\mylabel{S-TABLE6}\\
\dfrac{\forall i \in 1..n \; \exists j \in 1..n \;\;\;
       \senv \vdash k_{j} \subtype k_{i}' \;\;\;
       \senv \vdash k_{i}' \subtype k_{j} \;\;\;
       \senv \vdash v_{j} \subtype_{c} v_{i}'}
      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{closed} \subtype \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{closed}}
\end{array}
\]

In the next section we will show in more detail how our type system
uses \emph{unique}, \emph{open}, and \emph{closed} table types to
handle the refinement of table types.

@@

As an example, the rule \textsc{S-UNFOLDR} allows our type system to
type check the function \texttt{insert} from Section \ref{sec:alias}:
\begin{verbatim}
    local function insert (e:Element?, v:integer):Element
      return { info = v, next = e }
    end
\end{verbatim}
that is, the type checker uses the rule \textsc{S-UNFOLDR} to verify whether
the type of the table constructor is a subtype of \texttt{Element}:
\begin{align*}
\{ & ``info":\Integer, \\
   & ``next":\mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{regular} \cup \Nil \}_{unique} \subtype \\
& \mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{regular}
\end{align*}

@@

Our reduced core also splits function application, method application, and
the vararg expression (${...}$) into different syntactic categories.
It uses $e(el)_{s}$ to denote function applications that produce no value
because they appear as statements,
$e(el)_{e}$ to denote function applications that produce only one value,
even if the function application returns multiple values,
and $e(el)_{m}$ to denote function applications that produces as many values
as the function application returns.
It uses the same categories for method applications.
For vararg expressions, it only uses ${...}_{e}$ and ${...}_{m}$,
as the vararg expression cannot appear as a statement.

We also include two kinds of type coercions in our core language:
the expression ${<}t{>} \;n$ and the left-hand value $n[c] \; {<}t{>}$.
Both allow the refinement of table types.
We also split variable names into two categories to have safe aliasing
of tables in the presence of refinement.
We use $n_{e}$ when they appear as expressions and $n_{l}$ when they
appear as left-hand values.

We present the typing rules as a deduction system for the typing relations
$\env_{1}, \penv \vdash s, \env_{2}$ and $\env_{1}, \penv \vdash e : t, \env_{2}$.
We use the first one for typing statements and the second one for typing expressions.
The first relation means that given a type environment $\env_{1}$
and a projection environment $\penv$, typing a statement $s$ produces
a new type environment $\env_{2}$.
The second relation means that given a type environment $\env_{1}$
and a projection environment $\penv$, typing a expression $e$ has
type $t$ and produces a new type environment $\env_{2}$.
We need two typing relations because only expressions have types,
though expressions and statements can modify a type environment.

Even though we can assign only first-level types to variables,
our type system should be able to project unions of first-level types
from unions of second-level types.
Thus, we need two environments because $\env$ maps variables to first-level types,
while $\penv$ maps variables to second-level types.
We will show how our type system uses the environment $\penv$ for
handling projection types.
We use $\env_{1}[n \mapsto t]$ to extend the environment $\env_{1}$
with the variable $n$ that maps to type $t$.
We use $\penv[x \mapsto r]$ to extend the environment $\penv$ with
the variable $x$ that maps to the second-level type $r$.

\subsection{Assignment and function application}

Lua has multiple assignment, and the rule \textsc{T-ASSIGNMENT} uses
tuple types along with consistent-subtyping to type check this Lua feature:
\[
\begin{array}{c}
\mylabel{T-ASSIGNMENT}\\
\begin{array}{c}
\dfrac{\env_{1}, \penv \vdash el:r_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash \vec{l}:r_{2}, \env_{3} \;\;\;
       r_{1} \lesssim r_{2}}
      {\env_{1}, \penv \vdash \vec{l} = el,\env_{3}}
\end{array}
\end{array}
\]

As an example,
\[
x, y = 1, ``foo"
\]
type checks because
$1 \times ``foo" \times \Nil{*} \lesssim \Integer \times \String \times \Value{*}$,
assuming that $x$ and $y$ are in the environment with
types $\Integer$ and $\String$, respectively.

Our type system uses the type $\Value{*}$ to drop extra values
and the type $\Nil{*}$ to replace missing values.
The rule that type checks left-hand values list introduces the type
$\Value{*}$ to let the right-hand side produce more values than
expected in the left-hand side.
The rules that type check expression list introduce the type
$\Nil{*}$ to let the right-hand side produce fewer values than
expected in the left-hand side.
Our type system includes $\Nil{*}$ in the type of an expression list only
if its type does not end in another variadic type $t{*}$.
In the previous example, the rules \textsc{T-LHSLIST} and \textsc{T-EXPLIST2}
are the rules that introduce these types.
We define the rule \textsc{T-LHSLIST} as follows:
\[
\begin{array}{c}
\mylabel{T-LHSLIST}\\
\dfrac{\env, \penv \vdash l_{i}:t_{i}, \env_{i} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
       n = |\;\vec{l}\;|}
      {\env, \penv \vdash \vec{l}:t_{1} \times ... \times t_{n} \times \Value{*}, \env_{m}}
\end{array}
\]

First, this rule uses the same environment $\env$ to type check each
left-hand side value $l_{i}$.
Then, it takes each type $t_{i}$ and builds the tuple type for
the assignment rule, placing $\Value{*}$ in the end.
Table refinement can change the type environment while typing each left-hand value.
Thus, we use a partial auxiliary function \emph{merge} to collect
all the changes in a new environment $\env_{m}$, if there are no conflicts.

The rule \textsc{T-EXPLIST2} is similar to \textsc{T-LHSLIST}:
\[
\begin{array}{c}
\mylabel{T-EXPLIST2}\\
\dfrac{\env, \penv \vdash e_{i}:t_{i}, \env_{i} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
       n = |\;\vec{e}\;|}
      {\env, \penv \vdash \vec{e}:t_{1} \times ... \times t_{n} \times \Nil{*}, \env_{m}}
\end{array}
\]
